---
title: "spatial-intro"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, include = T, warning = F)

#A function to install and load missing packages

packages <- c("tidyverse", "sf", "stars", "raster", "here","rgdal", "lwgeom", "gstat",
              "ncmeta")

installLoadPackages <- function(pkg){
    new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
    if (length(new.pkg)) 
        install.packages(new.pkg, dependencies = TRUE,repos='http://cran.us.r-project.org')
    sapply(pkg, require, character.only = TRUE)
}
installLoadPackages(packages)

#You'll also need the patchwork package from github!!

# devtools::install_github("thomasp85/patchwork")
# devtools::install_github("michaeldorman/nngeo")
library(patchwork)

#Set relative path to data directories here !!
epu.dir <- here::here("data","EPU_shapefile")
data.dir <- here::here("data")

```

## Data structures

### Check out the structure of an `sf` object. 
```{r view_structure}
nc <- st_read(system.file("shape/nc.shp", package="sf"), quiet = T)

head(nc)

# Note that nc is a collection of simple features, each of which is described by a row

```

### Now see what it looks like
```{r plot_nc}
plot(nc)
```

### Create the simplest simple feature geometry: a `POINT`
```{r make_points}

point_in_space <- c(1,2) #in XY space 

point_geometry <- st_point(point_in_space)


point_geometry
plot(point_geometry)
```

### Combine multiple points into a `MULTIPOINT` geometry

```{r make_multipoints}
points_in_space <- rbind(c(3,5), c(4,6))

multipoint_geometry <- st_multipoint(points_in_space)

multipoint_geometry
plot(multipoint_geometry)
```

### Create a `LINESTRING` using the same points

```{r make_linestring}
linestring_geometry <- st_linestring(points_in_space)

linestring_geometry
plot(linestring_geometry)
plot(multipoint_geometry, add = T)
```

1. Create and plot a `POLYGON` geometry type using the function `st_polygon`. 

* `st_polygon` accepts lists of matrices of coordinate pairs (see `?st_polygon`). 
* The first and last coordinate pairs must be identical to close the polygon. For example, `list(matrix(c(x1, y1, ..., x1, y1), ncol  = 2, brow = TRUE))`.

```{r build-polygon}

```


2. Create and plot a `GEOMETRYCOLLECTION` geometry type by combining some of the above examples into one.

```{r build-geocol}


```

## Working with `sf`

### Getting `tidy` with it

1. Use your `dplyr` skills to determine the average `AREA` and `PERIMETER` of North Carolina counties that had more than 500 births between 1974-1979 (`BIR74`).

```{r getting-tidy}

```

2. Plot the top 10 largest counties in terms of `AREA` in North Carolina.

```{r plot-largest}


```

### `ggplot2` ft. `sf`

3. Using `ggplot2`, visualize the number of infant births by county between 1974-78 (`BIR74`). Next, visualize how birth rates changed by county between 1974-1978 and 1979-1984 (`BIR79`). 

```{r ggplot-sf}


```


### Reading spatial data

```{r epu_shapes}

#Read in shapefile using sf
sf_shape <- 
  st_read(file.path(epu.dir,"EPU_NOESTUARIES.shp"),quiet = T)

#Read in shapefile using rgdal
rgdal_shape <-
  readOGR(file.path(epu.dir,"EPU_NOESTUARIES.shp"),verbose = F)

epu <- sf_shape %>% dplyr::select(EPU)
plot(epu)
  
```

* Check out the class of the shapefile loaded with `rgdal`
```{r sp_to_sf}
#Read in shapefile using rgdal
class(rgdal_shape)
```

### Identifying existing Coordinate Reference System
```{r st-crs, echo = TRUE}
st_crs(epu)
```

#### Changing the Coordinate Reference System
```{r st-transform}

#Pass a proj4string
p4s <- "+proj=longlat +datum=NAD27 +no_defs"
new <- st_transform(epu, crs = st_crs(p4s))
new

#Pass an EPSG code
epsg <- 3857
new <- st_transform(epu, epsg)
```

4. The code below evaluates to a `SpatialPolygonsDataFrame`. Using `as()`, convert the output to an `sf` object. Assign the following `proj4string` to the object: "+proj=longlat +datum=WGS84 +no_defs". Plot the object using `ggplot2`.

```{r convert-polygon}
load(here::here("data/SpatialPolygonExample.rdata"))

```

### Calculating geometric distances

```{r beer, eval=T, echo = T}
#Wilmington is here
wilmy <- nc %>% filter(NAME == "New Hanover") %>% st_centroid() 

#Beer is here
wicked_weed <- st_geometry(st_point(c(-82.551440,35.591738)))
st_crs(wicked_weed) <- st_crs(nc)

#Draw a line between beer and Wilmington
a_long_ways_to_beer <- st_geometry( sf::st_linestring(
  matrix(rbind(st_coordinates(wilmy), st_coordinates(wicked_weed)), ncol = 2)) )
st_crs(a_long_ways_to_beer) <- st_crs(nc)

#Calculate the total distance with st_distance
total_dist <- st_distance(wilmy, wicked_weed)

ggplot() +
  geom_sf(data = nc) +
  geom_sf(data = wicked_weed) +
  geom_sf(data = wilmy) +
  geom_sf(data = a_long_ways_to_beer, color = "red") +
  geom_sf_text(data = st_centroid(a_long_ways_to_beer), 
               label = paste(round(total_dist,2), "m"), nudge_y = -0.2)
```

5. `st_distance` is just one of multiple `sf` geometric measurement functions, which also include `st_area` and `st_length`. Using `st_area`, calculate the area of the Mid-Atlantic Bight polygon in the `sf_shape` object.

```{r point-distance}

```


### Cropping sf objects

#### With a bounding box

6. Crop the `topography` `sf` object to the bounding box defined by `ymin`, `ymax`, `xmin`. and `xmax`. 

```{r otis_topo, eval=T, echo = T}

load(file.path(data.dir, "topo_4ft.Rdata"))

#Define a bounding box
ymax <-  41.6525; xmin <- -70.5975
ymin <-  41.6425; xmax <- -70.59


```


#### With an arbitary polygon

7. Use the `sf` polygon that you converted from `SpatialPolygons` class to crop the `topography` `sf` object. Try `st_intersection` and `st_difference`.

```{r ind_diff, eval = T, echo = T}


```

### Spatial Joins

8. Read in the shapefile `EPU_NOESTUARIES.shp`, and use `st_union` to join all polygons together. Next, try merging 2/4 polygons in the file using `dplyr` syntax. Visualize the results of boths spatial joins using `ggplot2`.

```{r join_epu, eval=T, echo = T}


```

### Putting it all together: An example with fake fish data

```{r fake_fishies}
load(file = file.path(data.dir, "catch_data.rdata"))

#Turn a data.frame into an sf object
catch_sf <- st_as_sf(catch, coords = c("X","Y"), crs = st_crs(sf_shape)) 

#Get length quantiles for our catch data
quantiles <- quantile(catch_sf$length, probs = seq(0,1,0.2))

#Join points into shapefile based on where they fall in the polygon
catch_sf <- st_join(catch_sf, sf_shape) %>% 
  mutate(quantiles = dplyr::case_when(length <= quantiles[2] ~ "first",
                                     length > quantiles[2] & length <= quantiles[3] ~ "second",
                                     length > quantiles[3] & length <= quantiles[4] ~ "third",
                                     length > quantiles[4] & length <= quantiles[5] ~ "fourth",
                                     length > quantiles[5] ~ "fifth"))

catch_sf$quantiles <- factor(catch_sf$quantiles,
                             levels = c("first","second","third","fourth","fifth"))

#Calculate quantiles for length data
ggplot() + 
  geom_sf(data = sf_shape) +
  geom_sf(data = catch_sf, aes(color = quantiles), show.legend = "point") +
  guides(color = guide_legend(title = "Length Quantiles")) +
  theme(legend.position='bottom')
```

#### Interpolating fake fish lengths with `gstat`

```{r NNI_p1, echo = T}

# Create a raster grid to interpolate over
r1 <- raster::raster()
e <- raster::extent(-76.83333, -65.66667, 34, 44.5)
raster::extent(r1) <- e
r1[] <- 1
raster::crs(r1) <- raster::crs(catch_sf)

#Convert our object to a SpatialPolygon and then into a raster. 
spdf <- sf_shape %>%
  st_union() %>% #Union all polygons
  st_transform(st_crs(catch_sf)) %>% #Make sure it has the same CRS as the fake fish data.
  as_Spatial()

polygon_rast <- raster::rasterize(spdf, r1) #Rasterize to mask output

#Convert our fake lengths sf object to SpatialPointsDataFrame for model input
catch_spatial_points <-
  catch_sf %>% 
  as_Spatial()

# Fit a model to estimate fish length at unsampled locations
# When nmax = 1, the model is a nearest neighbor model: predicted cells take on the value of their nearest neighbor
# When nmax > 1 and idp = 0, predictions are the mean of the two nearest neighbors w/o distance penalty.
# When idp > 0, a distance penalty of order idp is applied for a weighted mean

spatial_mod <- gstat::gstat(formula = length~1,
                            locations = catch_spatial_points, 
                            nmax = 5, set = list(idp = 0))


#Predict over the raster space
nn <- raster::interpolate(r1, spatial_mod)

#Check out the model output
plot(nn)
plot(catch_spatial_points, add = T)
```

```{r NNI_p2, echo = T}
#Mask the raster according to the original polygon extents
masked_model_output <- mask(nn, polygon_rast)

stars_output <- stars::st_as_stars(masked_model_output)

ggplot() +
  geom_stars(data = stars_output) +
  geom_sf(data = sf_shape, aes(color = EPU), fill = "transparent") 
```


### Your turn!!

9. Convert the following `data.frame` into an `sf` object. Pick a few variables and map them through time using `dplyr` and `ggplot2`. How do the variables differ by depth? And through time?

```{r EcoMon data}
load(file.path(data.dir, "LME_nutrients_spatial.Rdata"))
```

